|

### 2021-02-03 supplementary_information.Rmd
edits: si info, minor edits

  star_area <- stargazer::stargazer(m_area$fit[1],
                                    m_area$fit[2],
                                    m_area$fit[3],
                                    header = FALSE,
                                    type = "latex",
                                    covariate.labels = c("$\\sigma_{h}$",
                                                         "$\\sigma_{l}$",
                                                         "$\\sigma_{z}$",
                                                         "$\\phi$",
                                                         "$\\nu$",
                                                         "$\\alpha_{max}$",
                                                         "$\\theta$",
                                                         "$p_{d}$",
                                                         "Intercept"),
                                    single.row = FALSE,
                                    digits = 3,
                                    dep.var.caption  = "Response variable",
                                    dep.var.labels.include = FALSE,
                                    column.labels = c("Effect of $N_{p}$ on $\\alpha$ diversity",
                                                      "Effect of $N_{p}$ on $\\beta$ diversity",
                                                      "Effect of $N_{p}$ on $\\gamma$ diversity"),
                                    keep.stat = c("rsq"),
                                    model.numbers = FALSE)

@@

  star_bp <- stargazer::stargazer(m_bp$fit[1],
                                  m_bp$fit[2],
                                  m_bp$fit[3],
                                  header = FALSE,
                                  type = "latex",
                                  covariate.labels = c("$\\sigma_{h}$",
                                                       "$\\sigma_{l}$",
                                                       "$\\sigma_{z}$",
                                                       "$\\phi$",
                                                       "$\\nu$",
                                                       "$\\alpha_{max}$",
                                                       "$\\theta$",
                                                       "$p_{d}$",
                                                       "Intercept"),
                                  single.row = FALSE,
                                  digits = 3,
                                  dep.var.caption  = "Response variable",
                                  dep.var.labels.include = FALSE,
                                  column.labels = c("Effect of $P_{b}$ on $\\alpha$ diversity",
                                                    "Effect of $P_{b}$ on $\\beta$ diversity",
                                                    "Effect of $P_{b}$ on $\\gamma$ diversity"),
                                  keep.stat = c("rsq"),
                                  model.numbers = FALSE)

@@

We performed a sensitivity analysis of the metacommunity simulation to identify key simulation parameters that strongly affect the relationships between diversity metrics ($\alpha$, $\beta$, and $\gamma$) and ecosystem properties (the number of habitat patches $N_{p}$ and branching probability $P_{b}$). We generated 500 sets of parameter combinations by randomly drawing values of 8 simulation parameters from uniform distributions (**Table S1**). For each parameter combination, we generated 100 branching networks with the gradients of ecosystem size ($N_p \sim Unif(10, 150)$) and complexity ($P_b \sim Unif(0.01, 0.99)$). This results in a total of 50000 simulation replicates. In each simulation replicate, we allowed interspecific variation in niche optimum $\mu_i$ and width $\sigma_{niche,i}$ ($\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$, respectively; subscript $i$ represents species) and ran 1400 time steps of metacommunity dynamics. We obtained temporal means of $\alpha$, $\beta$, and $\gamma$ diversity for the last 1000 time steps. The first 400 time steps were discarded as initialization and burn-in periods.

@@

where $\psi_q$ ($q = 0-2$) are the intercept ($\psi_0$) and regression coefficients ($\psi_1$ and $\psi_2$). We extracted 500 estimates of $\psi_1$ and $\psi_2$, which represent the effects of $N_{p}$ and $P_{b}$ on diversity metrics under a given parameter combination. To examine influences of simulation parameters (**Table S1**) on $\psi_1$ and $\psi_2$, we developed the following regression model taking $\psi_1$ or $\psi_2$ as a response variable $u_k$ (parameter combination $k$):

$$
\begin{aligned}
u_{k} &\sim Normal(\mu_k, \sigma^2)\\
\mu_k &= \zeta_0 + \zeta_1 \sigma_{h,k} + \zeta_2 \sigma_{l,k} + \zeta_3 \sigma_{z,k} + \zeta_4 \phi_k + \zeta_5 \nu_k + \zeta_6 \alpha_{max, k} + \zeta_7 \theta_k + \zeta_8 p_{d,k} 
\end{aligned}
$$
where $\zeta_q$ ($q = 0-8$) are the intercept and regression coefficients. Explanatory variables were standardized to a mean of zero and a standard deviation of one, so that regression coefficients are comparable.

@@

Longitudinal gradients of local species richness have been extensively studied in rivers, illuminating typical patterns observed in nature. The most common pattern is a downstream increase of local species richness[@mcdowallDiadromyAssemblyRestoration1996; @mcdowallFightingFlowDownstreamupstream1998; @rahelFishAssemblagesHabitat1991; @teruiThreeEcologicalFactors2016]. However, recent empirical and theoretical studies also showed 'reversed patterns,' in which local species richness decreases downstream[@bsemerHeadwatersAreCritical2013; @harveyDisturbanceReversesClassic2018]. We predicted the longitudinal gradient of local species richness to confirm that our simulation scenarios are capable of reproducing the previously observed patterns of local species richness. We considered 32 simulation scenarios comprising four landscape and eight ecological scenarios, as described in the main text (a set of parameters is described in **Table S2**). Under each simulation scenario, we generated 10 branching networks with fixed parameters of ecosystem size ($N_p = 100$) and complexity ($P_b = 0.5$). This results in a total of 320 simulation replicates. In each simulation replicate, we allowed interspecific variation in niche optimum $\mu_i$ and width $\sigma_{niche,i}$ ($\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$, respectively; subscript $i$ represents species) and ran 1400 time steps of metacommunity dynamics. We obtained temporal means of local species richness at each habitat patch for the last 1000 time steps. The first 400 time steps were discarded as initialization and burn-in periods. We evaluated the relationship between local species richness and the number of upstream habitat patches, a proxy for the longitudinal position of a habitat patch.

@@

**Midwest, US.** We assembled fish community data collected by the Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. These data sets cover most of Upper Mississippi (Hydrologic Unit Code 2 [HUC 2] , region 07, as defined
by U.S. Geological Survey and U.S. Department of Agriculture Natural Resources Conservation Service[@u.s.geologicalsurveyandu.s.departmentofagriculturenaturalresourcesconservationserviceFederalStandardsProcedures2013]) and the part of Great Lakes (HUC 2, region 04), Missouri (HUC 2, region 10), and Ohio (HUC 2, region 05). Fish data were collected from summer to fall with electrofishing (backpack, barge-type, or boat-mounted) and supplemental netting at some locations. We screened data through the following procedure:

@@

We evaluated sensitivity of asymptotic species richness (Chao 2 estimator) to sample size (i.e., the number of sampling sites in a watershed). We simulated presence-absence data of species with a known value of true species richness $S_{true}$ and the number of sampling sites $N_{site}$. In this simulation, presence of species $i$ at site $x$, $I_{ix}$, was drawn from a Bernoulli distribution as $I_{ix} \sim Bernoulli(p_{i}\kappa_{ix})$ where $p_i$ is the detection probability for species $i$ and $\kappa_{ix}$ is the presence probability of species $i$ at site $x$. Based on the incidence frequency of simulated species $F_i = \sum_{x = 1}^{N_{site}}I_{ix}$, we estimated asymptotic species richness using the iNEXT function in the R package 'iNEXT'[@hsiehINEXTPackageRarefaction2016]. We calculated % bias of estimated asymptotic species richness $S_{est}$:

@@

The number of sampling sites influenced the estimation accuracy of asymptotic species richness. The % bias decreased sharply as the number of sampling sites increased (**Figure S20**). In particular, the estimation bias with a small sample size ($N_{site} = 5$) was substantial when the true species richness $S_{true}$ was low; some estimates showed > 150 % bias (**Figure S20**). Given the simulation results, estimates of asymptotic species richness at watersheds with < 10 sampling sites may involve substantial statistical uncertainty.

@@

  figure_cap[i] <- paste0("Theoretical predictions for longitudinal gradients of local species richness in branching networks. The longitudinal position of each habitat patch (x axis) was expressed as the number of upstream contributing patches. `r conditional_text[",i,"]` Lines and shades are loess curves fitted to simulated data and their 95% confidence intervals. Each panel represents different ecological scenarios under which metacommunity dynamics were simulated. Rows represent different competition strength. Competitive coefficients ($\\alpha_{ij}$) were varied randomly from 0 to 1.5 (top, strong competition) or 0.75 (bottom, weak competition). Columns and lines represent different dispersal scenarios (dispersal distance and probability). Left and right columns show long-distance (the rate parameter of an exponential dispersal kernel $\\theta = 0.10$) and short-distance dispersal ($\\theta = 1.0$) scenarios repectively. Red and blue lines show low ($p_d = 0.01$) and high dispersal probabilities ($p_d = 0.10$). Other parameters are as follows: environmental variation at headwaters $\\sigma_{h} = `r param$sigma_h[",i,"]`$; local environmental noise $\\sigma_{l} = `r param$sigma_l[",i,"]`$; ecosystem size $N_{p} = 100$; ecosystem complexity $P_{b} = 0.5$.")
}

@@

**Figure S20 ** Sensitivity analysis of asymptotic species richness in relation to true species richness $S_{true}$ (panels) and the number of sampling sites $N_{site}$ (x-axis). Positive values of % bias indicate an overestimation of species richness, while negative values indicate an underestimation. Different panels show results with different true species richness. The box boundaries represent 25 and 75 percentiles and the whiskers represent 5 and 95 percentiles. Dots represent individual simulation replicates.

>>>>

  star_area <- stargazer::stargazer(m_area$fit[1],
                                    m_area$fit[2],
                                    m_area$fit[3],
                                    header = FALSE,
                                    type = "latex",
                                    covariate.labels = c("$\\sigma_{h}$",
                                                         "$\\sigma_{l}$",
                                                         "$\\sigma_{z}$",
                                                         "$\\phi$",
                                                         "$\\nu$",
                                                         "$\\alpha_{max}$",
                                                         "$\\theta$",
                                                         "$p_{d}$",
                                                         "Intercept"),
                                    single.row = FALSE,
                                    digits = 3,
                                    dep.var.caption  = "Response variable",
                                    dep.var.labels.include = FALSE,
                                    column.labels = c("Effect of $N_{p}$ on $\\alpha$ diversity",
                                                      "Effect of $N_{p}$ on $\\beta$ diversity",
                                                      "Effect of $N_{p}$ on $\\gamma$ diversity"),
                                    report = "vcs",
                                    omit.table.layout = "ns",
                                    model.numbers = FALSE)

@@

  star_bp <- stargazer::stargazer(m_bp$fit[1],
                                  m_bp$fit[2],
                                  m_bp$fit[3],
                                  header = FALSE,
                                  type = "latex",
                                  covariate.labels = c("$\\sigma_{h}$",
                                                       "$\\sigma_{l}$",
                                                       "$\\sigma_{z}$",
                                                       "$\\phi$",
                                                       "$\\nu$",
                                                       "$\\alpha_{max}$",
                                                       "$\\theta$",
                                                       "$p_{d}$",
                                                       "Intercept"),
                                  single.row = FALSE,
                                  digits = 3,
                                  dep.var.caption  = "Response variable",
                                  dep.var.labels.include = FALSE,
                                  column.labels = c("Effect of $P_{b}$ on $\\alpha$ diversity",
                                                    "Effect of $P_{b}$ on $\\beta$ diversity",
                                                    "Effect of $P_{b}$ on $\\gamma$ diversity"),
                                  report = "vcs",
                                  omit.table.layout = "ns",
                                  model.numbers = FALSE)

@@

We performed a sensitivity analysis of the metacommunity simulation to identify key simulation parameters that strongly affect the relationships between diversity metrics ($\alpha$, $\beta$, and $\gamma$) and ecosystem properties (the number of habitat patches $N_{p}$ and branching probability $P_{b}$). We generated 500 sets of parameter combinations by randomly drawing values of 8 simulation parameters from uniform distributions (**Table S1**). For each parameter combination, we generated 100 branching networks with the gradients of ecosystem size ($N_p \sim Unif(10, 150)$) and complexity ($P_b \sim Unif(0.01, 0.99)$). This results in a total of 50000 simulation replicates. In each simulation replicate, we allowed interspecific variation in niche optimum $\mu_i$ and width $\sigma_{niche,i}$ ($\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$, respectively; subscript $i$ represents species) and ran 1400 time steps of metacommunity dynamics. We obtained temporal means of $\alpha$, $\beta$, and $\gamma$ diversity for the last 1000 time steps. The first 400 time steps were discarded as initialization and burn-in periods. We removed simulation replicates in which no species established populations over the initial 400 time steps.

@@

where $\psi_k$ ($k = 0-2$) are the intercept ($\psi_0$) and regression coefficients ($\psi_1$ and $\psi_2$). We extracted 500 estimates of $\psi_1$ and $\psi_2$, which represent the effects of $N_{p}$ and $P_{b}$ on diversity metrics under a given parameter combination. To examine influences of simulation parameters (**Table S1**) on $\psi_1$ and $\psi_2$, we developed the following regression model taking $\psi_1$ or $\psi_2$ as a response variable $u_n$ (parameter combination $n$):

$$
\begin{aligned}
u_{n} &\sim Normal(\mu_n, \sigma^2)\\
\mu_n &= \zeta_0 + \zeta_1 \sigma_{h,n} + \zeta_2 \sigma_{l,n} + \zeta_3 \sigma_{z,n} + \zeta_4 \phi_n + \zeta_5 \nu_n + \zeta_6 \alpha_{max, n} + \zeta_7 \theta_n + \zeta_8 p_{d,n} 
\end{aligned}
$$
where $\zeta_k$ ($k = 0-8$) are the intercept ($\zeta_0$) and regression coefficients ($\zeta_{1-8}$). Explanatory variables were standardized to a mean of zero and a standard deviation of one, so that regression coefficients are comparable.

@@

Longitudinal gradients of local species richness have been extensively studied in rivers, illuminating typical patterns observed in nature. The most common pattern is a downstream increase of local species richness[@mcdowallDiadromyAssemblyRestoration1996; @mcdowallFightingFlowDownstreamupstream1998; @rahelFishAssemblagesHabitat1991; @teruiThreeEcologicalFactors2016]. However, recent empirical and theoretical studies also showed 'reversed patterns,' in which local species richness decreases downstream[@bsemerHeadwatersAreCritical2013; @harveyDisturbanceReversesClassic2018]. We predicted the longitudinal gradient of local species richness to confirm that our simulation scenarios are capable of reproducing the previously observed patterns of local species richness. We considered 32 simulation scenarios comprising four landscape and eight ecological scenarios, as described in the main text (a set of parameters is described in **Table S2**). Under each simulation scenario, we generated 10 branching networks with fixed parameters of ecosystem size ($N_p = 100$) and complexity ($P_b = 0.5$). This results in a total of 320 simulation replicates. In each simulation replicate, we allowed interspecific variation in niche optimum $\mu_i$ and width $\sigma_{niche,i}$ ($\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$, respectively; subscript $i$ represents species) and ran 1400 time steps of metacommunity dynamics. We obtained temporal means of local species richness at each habitat patch for the last 1000 time steps. The first 400 time steps were discarded as initialization and burn-in periods. We removed simulation replicates in which no species established populations over the initial 400 time steps. We evaluated the relationship between local species richness and the number of upstream habitat patches, a proxy for the longitudinal position of a habitat patch.

@@

**Midwest, US.** We assembled fish community data collected by the Iowa Department of Natural Resources[@wiltonBiologicalAssessmentIowa2004], Illinois Environmental Protection Agency and Illinois Department of Natural Resources[@ilepaIllinoisWaterMonitoring2014], Minnesota Pollution Control Agency[@mpcaDevelopmentFishbasedIndex2014], and Wisconsin Department of Natural Resources[@wdnrGuidelinesAssessingFish2018]. These data sets cover most of Upper Mississippi (Hydrologic Unit Code 2 [HUC 2] , region 07, as defined
by U.S. Geological Survey and U.S. Department of Agriculture Natural Resources Conservation Service[@u.s.geologicalsurveyandu.s.departmentofagriculturenaturalresourcesconservationserviceFederalStandardsProcedures2013]) and the part of Great Lakes (HUC 2, region 04), Missouri (HUC 2, region 10), and Ohio (HUC 2, region 05). Fish data were collected from summer to fall with electrofishing (backpack, barge-type, or boat-mounted) and supplemental netting at some locations. We screened data through the following procedure:

@@

We evaluated sensitivity of asymptotic species richness (Chao 2 estimator) to sample size (i.e., the number of sampling sites in a watershed). We simulated presence-absence data of species with known values of true species richness $S_{true}$ and the number of sampling sites $N_{site}$. In this simulation, presence of species $i$ at site $x$, $J_{ix}$, was drawn from a Bernoulli distribution as $J_{ix} \sim Bernoulli(p_{i}\kappa_{ix})$ where $p_i$ is the detection probability for species $i$ and $\kappa_{ix}$ is the presence probability of species $i$ at site $x$. Based on the incidence frequency of simulated species $F_i = \sum_{x = 1}^{N_{site}}J_{ix}$, we estimated asymptotic species richness using the iNEXT function in the R package 'iNEXT'[@hsiehINEXTPackageRarefaction2016]. We calculated % bias of estimated asymptotic species richness $S_{est}$:

@@

The number of sampling sites influenced the estimation accuracy of asymptotic species richness. The % bias decreased sharply as the number of sampling sites increased (**Figure S20**). In particular, the estimation bias with a small sample size ($N_{site} = 5$) was substantial when the true species richness $S_{true}$ was low; some estimates showed > 150% bias (**Figure S20**). Given the simulation results, estimates of asymptotic species richness at watersheds with < 10 sampling sites may involve substantial statistical uncertainty.

@@

  figure_cap[i] <- paste0("Theoretical predictions for longitudinal gradients of local species richness in branching networks. The longitudinal position of each habitat patch (x-axis) was expressed as the number of upstream habitat patches. `r conditional_text[",i,"]` Lines and shades are loess curves fitted to simulated data and their 95% confidence intervals. Each panel represents different ecological scenarios under which metacommunity dynamics were simulated. Rows represent different competition strength. Competitive coefficients ($\\alpha_{ij}$) were varied randomly from 0 to 1.5 (top, strong competition) or 0.75 (bottom, weak competition). Columns and lines represent different dispersal scenarios (dispersal distance and probability). Left and right columns show long-distance (the rate parameter of an exponential dispersal kernel $\\theta = 0.10$) and short-distance dispersal ($\\theta = 1.0$) scenarios repectively. Red and blue lines show low ($p_d = 0.01$) and high dispersal probabilities ($p_d = 0.10$). Other parameters are as follows: environmental variation at headwaters $\\sigma_{h} = `r param$sigma_h[",i,"]`$; local environmental noise $\\sigma_{l} = `r param$sigma_l[",i,"]`$; ecosystem size $N_{p} = 100$; ecosystem complexity $P_{b} = 0.5$.")
}

@@

**Figure S20 ** Sensitivity analysis of asymptotic species richness in relation to true species richness $S_{true}$ (panels) and the number of sampling sites $N_{site}$ (x-axis). Positive values of % bias indicate an overestimation of species richness, while negative values indicate an underestimation. Different panels show results with different true species richness. The center lines are median values, the box boundaries 25 and 75 percentiles, and the whiskers 5 and 95 percentiles. Dots represent individual simulation replicates.
